---
title: "Predicting Online News Article Popularity"
author: "Sonam"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: true
    highlight: tango
    toc: 3
    toc_float: true
    code_folding: hide
    fig_width: 8 
    fig_height: 5
---
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require('pacman'))install.packages('pacman')
pacman::p_load(DT, miscTools, caret, ROCR, pROC, e1071, C50,
               rpart,rpart.plot, rattle, RColorBrewer, ggplot2,
               dplyr,randomForest)
```

# Introduction

The aim of this projecr is to explore a dataset in depth, apply a business analytics mindset to implement appropriate predictive analytics, and communicate the findings effectively.

The dataset is comprised of some statistical measures on online news articles. Hence, this analysis builds a machine learning system on the dataset to predict the popularity of online news articles. The goal of the analysis is to use this system to configure and present future articles that sell more advertisement.


# Dataset

The dataset set comes from [UCI Machine Learning Repository: Online News Popularity Data Set](https://archive.ics.uci.edu/ml/datasets/Online+News+Popularity). This dataset summarizes a heterogeneous set of features about articles published by Mashable in a period of two years. The goal is to predict the number of shares in social networks (popularity).

## Data Description

Attribute Information:  

     0. url:                           URL of the article
     1. timedelta:                     Days between the article publication and the dataset acquisition
     2. n_tokens_title:                Number of words in the title
     3. n_tokens_content:              Number of words in the content
     4. n_unique_tokens:               Rate of unique words in the content
     5. n_non_stop_words:              Rate of non-stop words in the content
     6. n_non_stop_unique_tokens:      Rate of unique non-stop words in the content
     7. num_hrefs:                     Number of links
     8. num_self_hrefs:                Number of links to other articles published by Mashable
     9. num_imgs:                      Number of images
    10. num_videos:                    Number of videos
    11. average_token_length:          Average length of the words in the content
    12. num_keywords:                  Number of keywords in the metadata
    13. data_channel_is_lifestyle:     Is data channel 'Lifestyle'?
    14. data_channel_is_entertainment: Is data channel 'Entertainment'?
    15. data_channel_is_bus:           Is data channel 'Business'?
    16. data_channel_is_socmed:        Is data channel 'Social Media'?
    17. data_channel_is_tech:          Is data channel 'Tech'?
    18. data_channel_is_world:         Is data channel 'World'?
    19. kw_min_min:                    Worst keyword (min. shares)
    20. kw_max_min:                    Worst keyword (max. shares)
    21. kw_avg_min:                    Worst keyword (avg. shares)
    22. kw_min_max:                    Best keyword (min. shares)
    23. kw_max_max:                    Best keyword (max. shares)
    24. kw_avg_max:                    Best keyword (avg. shares)
    25. kw_min_avg:                    Avg. keyword (min. shares)
    26. kw_max_avg:                    Avg. keyword (max. shares)
    27. kw_avg_avg:                    Avg. keyword (avg. shares)
    28. self_reference_min_shares:     Min. shares of referenced articles in Mashable
    29. self_reference_max_shares:     Max. shares of referenced articles in Mashable
    30. self_reference_avg_sharess:    Avg. shares of referenced articles in Mashable
    31. weekday_is_monday:             Was the article published on a Monday?
    32. weekday_is_tuesday:            Was the article published on a Tuesday?
    33. weekday_is_wednesday:          Was the article published on a Wednesday?
    34. weekday_is_thursday:           Was the article published on a Thursday?
    35. weekday_is_friday:             Was the article published on a Friday?
    36. weekday_is_saturday:           Was the article published on a Saturday?
    37. weekday_is_sunday:             Was the article published on a Sunday?
    38. is_weekend:                    Was the article published on the weekend?
    39. LDA_00:                        Closeness to LDA topic 0
    40. LDA_01:                        Closeness to LDA topic 1
    41. LDA_02:                        Closeness to LDA topic 2
    42. LDA_03:                        Closeness to LDA topic 3
    43. LDA_04:                        Closeness to LDA topic 4
    44. global_subjectivity:           Text subjectivity
    45. global_sentiment_polarity:     Text sentiment polarity
    46. global_rate_positive_words:    Rate of positive words in the content
    47. global_rate_negative_words:    Rate of negative words in the content
    48. rate_positive_words:           Rate of positive words among non-neutral tokens
    49. rate_negative_words:           Rate of negative words among non-neutral tokens
    50. avg_positive_polarity:         Avg. polarity of positive words
    51. min_positive_polarity:         Min. polarity of positive words
    52. max_positive_polarity:         Max. polarity of positive words
    53. avg_negative_polarity:         Avg. polarity of negative  words
    54. min_negative_polarity:         Min. polarity of negative  words
    55. max_negative_polarity:         Max. polarity of negative  words
    56. title_subjectivity:            Title subjectivity
    57. title_sentiment_polarity:      Title polarity
    58. abs_title_subjectivity:        Absolute subjectivity level
    59. abs_title_sentiment_polarity:  Absolute polarity level
    60. shares:                        Number of shares (target)



# Importing the data 

The data is imported into R and it comprises of 61 features.The 61st feature i.e. **shares** is the target variable i.e. an article is eligible for publishing or good enough to sell advertisement if the number of shares are more than 14000.


```{r}
news=read.csv("OnlineNewsPopularity/OnlineNewsPopularity.csv", stringsAsFactors = F)
names(news)
```

# Explore and Clean the dataset

To find the which of the remaining 60 features are best for predicting the **shares** we must inspect and look for patterns in the data through summaries and plots.

## Exploration: Distributions of features 


Lets check the distribution of values that each column takes.
To do so we need to filter out the columns that have numeric values only and as seen from the result of `str(news)` below all the columns except `url` are numeric.

```{r}
str(news)
```

With `url` removed we also check if there are any nulls or NA in the data , turns out there arent any NULLs in the data. 

```{r}
is.null(news)

par(mfrow= c(3,4))
for(i in 2:length(news)){
  hist(news[,i],main  = names(news)[i] ,xlab=names(news)[i])
}
```


## Cleaning: Handle Categorical Data

Looking at the distributions of  all the columns above we can conclude that all the columns 14-19 and 32-39 (see the names below) have binary data, hence it would be better to convert these into a factor.

```{r}
names(news)[c(14:19,32:39)]
```

Converting the columns mentioned above to factors.

```{r}
for (i in names(news)[c(14:19,32:39)]){
  news[,i]<-factor(news[,i])
}
str(news[,names(news)[c(14:19,32:39)]])
```



## Cleaning: Handle Missing Data

Missing values are coded as 0 in this dataset.Keeping apart the binary data columns there are around 3000 rows with missing data that needs to be cleaned first.


Columns with missing data:

```{r}
names(news)[c(11,20,44,45,46,48,49,50,53)]
```

Total records vs no. of unclean records

```{r}
for(i in c(11,20,44,45,46,48,49,50,53))news_clean <- news[news[,i]!=0,]
print(paste0("Total Records: ",nrow(news), 
             " Unclean Records: ", nrow(news)-nrow(news_clean),
             " i.e. ", 100*(nrow(news)-nrow(news_clean))/nrow(news), ' %' ))
```


Since the number of unclean rows is close to 3% , hence we can omit the bad rows.

```{r}
news <- news_clean
```



## Transforming: Handle Skewness in Distributions

Some of the variables have heavily right skewed distributions, including the response `shares`. So we will transform them to reduce the skewness. For those variables with all values bigger than 0, we use log, and other variable with 0, we use square root to transform them.We are omitting the response `shares` here.

Columns undergoing transformation:

```{r}
names(news)[c(3,7,8,9,10,27:30,40:43,47)]
```


```{r}
backup<-news

for(i in c(3,7,8,9,10,27:30,40:43,47)){
  if(!sum(news[,i]==0)){
    news[,i] <- log(news[,i]); 
    #names(news)[i] <- paste("log_",names(news)[i], sep="")
  }
  else{
    news[,i] <- sqrt(news[,i]); 
    #names(news)[i] <- paste("sqrt_",names(news)[i], sep="")
    }
}


DT::datatable(head(news,3))
```


## Exploration: Number of shares by the weekday

Observing the pattern of number of shares on each week day, it was found that publishing day didn’t show much influence on shares. 
We created a new categorical column called as `news_day` using the `weekday_is_*` columns to get all the days in a single column. This helped in formatting the patterns in the plot more efficiently.

```{r}
news$news_day <- rep("Sunday", nrow(news))
news$news_day[news$weekday_is_monday==1] <- "Monday"
news$news_day[news$weekday_is_tuesday==1] <- "Tuesday"
news$news_day[news$weekday_is_wednesday==1] <- "Wednesday"
news$news_day[news$weekday_is_thursday==1] <- "Thursday"
news$news_day[news$weekday_is_friday==1] <- "Friday"
news$news_day[news$weekday_is_saturday==1] <- "Saturday"
news$news_day<-factor(news$news_day, levels = c('Sunday','Monday','Tuesday',
                                                'Wednesday','Thursday','Friday',
                                                'Saturday'))
#Check 
p1 <- ggplot(data=news, aes(news_day, log(shares)))
p1 + geom_boxplot()
news <- news[,-c(32:38,62)]

```

So we get rid of all the indicators but leave `is_weekend` because some difference is there bewtween weekdays and weekend data.

## Exploration: Number of shares by publish tags

Repeating the same process to find patterns in number of shares for different publish tags.

```{r}
news$data_channel <- rep("other", nrow(news))
news$data_channel[news$data_channel_is_lifestyle==1] <- "Lifestyle"
news$data_channel[news$data_channel_is_entertainment==1] <- "Entertainment"
news$data_channel[news$data_channel_is_bus==1] <- "Bus"
news$data_channel[news$data_channel_is_socmed==1] <- "Socmed"
news$data_channel[news$data_channel_is_tech==1] <- "Tech"
news$data_channel[news$data_channel_is_world==1] <- "World"
news$data_channel<-factor(news$data_channel)
#Check 
p1 <- ggplot(data=news, aes(data_channel, log(shares)))
p1 + geom_boxplot()
```

There are very mild variations in means with the data channels so we will keep these to see if they are significant.

## Cleaning: Remove unwanted Columns

Removing `datachannel` created for the plot above.
```{r}
news <- news[,-c(55)]
```

Also deleting the url and timedelta columns.

```{r}
news <- subset( news, select = -c(url, timedelta ) )
```

Deleting column `n_non_stop_words` since it has only one value , hence its a constant.

```{r}
news <- subset( news, select = -c(n_non_stop_words ) )
```

## Transformation: Generate binary response variable

Define articles with shares larger than 1400 (median) as popular article.

```{r}
news$shares <- factor(ifelse(news$shares>1400,1,0))
```

# Modeling


Since our target is a class i.e. 1 for a popular aricle and 0 for a non popular article , we will apply classification methods like Knn , Classification and Regression Trees,C5.0 Trees and Random Forests to train our data and predict on a test set. We will generate the training and test sets for the same.


## Create Training and Test sets

Split Train - 70%; Test - 30% and set a color palette for each method.

```{r warning=FALSE,message=FALSE}
#set random situation
set.seed(100)
# Select traning data and prediction data
ind<-sample(2,nrow(news),replace=TRUE,prob=c(0.7,0.3))

color.knn<-'#efab69'
color.lda<-'#ab69ef'
color.qda<-'#69adef'
color.lr<-'#adef69'
color.cart<-"#72E2FF"
color.c50<-"#67B5DA"
color.rf<-"#68B518"
print(paste0('#Train: ', table(ind)[1],' #Test: ', table(ind)[2]))
```


## Check for collinearity 

We need to see if any  numerical columms are collinear to each before applying our algorithms. 

```{r}

corDF = cor(news[ind==1,names(dplyr::select_if(news, is.numeric))]);
dissimilarity <- 1 - abs(corDF);
distance <- as.dist(dissimilarity);
hc <- hclust(distance);  
clusterV = cutree(hc,h=0.05);
df<-as.data.frame(clusterV)
df$columns<-rownames(df)
knitr::kable(df[(df$clusterV==32),])
DT::datatable(df)
```



From the cluster 32 we can see that  `rate_positive_words` and `rate_negative_words`. Hence, we reomve `rate_negative_words`

```{r}
news<-news[,-c(40)]
DT::datatable(head(news))
```




## LDA

Train the model on train set

```{r}
news.lda<- train(shares ~., method='lda', data=news[ind==1,])
summary(news.lda)
```

Predict on test set 

```{r}
news.lda.pred <- predict( news.lda,news[ind==2,])
news.lda.prob <- predict(news.lda,news[ind==2,],type='prob')
```

Confusion matrix for test set

```{r}
confusionMatrix(news.lda.pred, news[ind==2,]$shares)
```

ROC Curve

```{r}
news.lda.roc <- roc(news[ind==2,]$shares,news.lda.prob[,2])
plot(news.lda.roc, print.auc=TRUE, auc.polygon=TRUE, grid=c(0.1, 0.2),
     grid.col=c("green", "red"), max.auc.polygon=TRUE,
     auc.polygon.col=color.lda, print.thres=TRUE)

```


## QDA 

Train the model on train set

```{r}
news.qda<- train(shares ~., method='qda', data=news[ind==1,])
summary(news.qda)
```

Predict on test set 

```{r}
news.qda.pred <- predict( news.qda,news[ind==2,])
news.qda.prob <- predict(news.qda,news[ind==2,],type='prob')
```

Confusion matrix for test set

```{r}
confusionMatrix(news.qda.pred, news[ind==2,]$shares)
```

ROC Curve

```{r}
news.qda.roc <- roc(news[ind==2,]$shares,news.qda.prob[,2])
plot(news.qda.roc, print.auc=TRUE, auc.polygon=TRUE, grid=c(0.1, 0.2),
     grid.col=c("green", "red"), max.auc.polygon=TRUE,
     auc.polygon.col=color.qda, print.thres=TRUE)
```

## Logistic Regression

Train the model on train set

```{r}

news.lr<- train(shares ~., method='glm',family='binomial', data=news[ind==1,])
summary(news.lr)
```

Predict on test set 

```{r}
news.lr.pred <- predict( news.lr,news[ind==2,])
news.lr.prob <- predict(news.lr,news[ind==2,],type='prob')
```

Confusion matrix for test set

```{r}
confusionMatrix(news.lr.pred, news[ind==2,]$shares)
```

ROC Curve

```{r}
news.lr.roc <- roc(news[ind==2,]$shares,news.lr.prob[,2])
plot(news.lr.roc, print.auc=TRUE, auc.polygon=TRUE, grid=c(0.1, 0.2),
     grid.col=c("green", "red"), max.auc.polygon=TRUE,
     auc.polygon.col=color.lr, print.thres=TRUE)
```

## KNN

Train the model on train set

```{r}
news.knn <- knn3(shares ~.,news[ind==1,])
summary(news.knn)
```

Predict on test set 

```{r}
news.knn.pred <- predict( news.knn,news[ind==2,],type='class')
news.knn.prob <- predict(news.knn,news[ind==2,],type='prob')
```

Confusion matrix for test set

```{r}
confusionMatrix(news.knn.pred, news[ind==2,]$shares)
```

ROC Curve

```{r}
news.knn.roc <- roc(news[ind==2,]$shares,news.knn.prob[,2])
plot(news.knn.roc, print.auc=TRUE, auc.polygon=TRUE, grid=c(0.1, 0.2),
     grid.col=c("green", "red"), max.auc.polygon=TRUE,
     auc.polygon.col=color.knn, print.thres=TRUE)

```

## CART

Train the model on train set

```{r}
news.cart<-rpart(shares ~.,news[ind==1,],method='class')
summary(news.cart)
```

Plot tree

```{r}
fancyRpartPlot(news.cart) 
```

Predict on test set 

```{r}
news.cart.pred<-predict( news.cart,news[ind==2,] ,type="class")
news.cart.prob<-predict( news.cart,news[ind==2,] ,type="prob")
```

Confusion matrix for test set

```{r}
confusionMatrix(news.cart.pred, news[ind==2,]$shares)
```

ROC Curve

```{r}
news.cart.roc <- roc(news[ind==2,]$shares,news.cart.prob[,2])
plot(news.cart.roc, print.auc=TRUE, auc.polygon=TRUE, grid=c(0.1, 0.2),
     grid.col=c("green", "red"), max.auc.polygon=TRUE,
     auc.polygon.col=color.cart, print.thres=TRUE)


```

## C5.0

Train the model on train set

```{r}
news.c50<-C5.0(shares ~.,news[ind==1,],method='class')
summary(news.c50)
```

Predict on test set 

```{r}
news.c50.pred<-predict( news.c50,news[ind==2,] ,type="class")
news.c50.prob<-predict( news.c50,news[ind==2,] ,type="prob")
```

Confusion matrix for test set

```{r}
confusionMatrix(news.c50.pred, news[ind==2,]$shares)
```

ROC Curve

```{r}
news.c50.roc <- roc(news[ind==2,]$shares,news.c50.prob[,2])
plot(news.c50.roc, print.auc=TRUE, auc.polygon=TRUE, grid=c(0.1, 0.2),
     grid.col=c("green", "red"), max.auc.polygon=TRUE,
     auc.polygon.col=color.cart, print.thres=TRUE)


```

## Random Forest

Train the model on train set

```{r}
news.rf<-randomForest(shares ~.,news[ind==1,],method='class')
summary(news.rf)
```

Plot number of trees vs error

```{r}

plot(news.rf)
```

Plot feature importance
 
 
```{r}
varImpPlot(news.rf)
```

Predict on test set 

```{r}
news.rf.pred<-predict( news.rf,news[ind==2,], type="class")
news.rf.prob<-predict( news.rf,news[ind==2,], type="prob")
```

Confusion matrix for test set

```{r}
confusionMatrix(news.rf.pred, news[ind==2,]$shares)
```

ROC Curve

```{r}
news.rf.roc <- roc(news[ind==2,]$shares,news.rf.prob[,2])
plot(news.rf.roc, print.auc=TRUE, auc.polygon=TRUE, grid=c(0.1, 0.2),
     grid.col=c("green", "red"), max.auc.polygon=TRUE,
     auc.polygon.col=color.cart, print.thres=TRUE)


```


## Model Comparison

```{r}
ROCCurve<-par(pty = "s")
plot(performance(prediction(news.knn.prob[,2],news[ind==2,]$shares),'tpr','fpr'),
     col=color.knn, lwd=3)
text(0.55,0.6,"KNN",col=color.knn)
plot(performance(prediction(news.cart.prob[,2],news[ind==2,]$shares),'tpr','fpr'),
     col=color.cart, lwd=3, add=TRUE)
text(0.3,0.4,"CART",col=color.cart)
plot(performance(prediction(news.c50.prob[,2],news[ind==2,]$shares),'tpr','fpr'),
     col=color.c50, lwd=3, add=TRUE)
text(0.15,0.5,"C5.0",col=color.c50)
plot(performance(prediction(news.rf.prob[,2],news[ind==2,]$shares),'tpr','fpr'),
     col=color.rf, lwd=3, add=TRUE)
text(0.3,0.7,"Random Forest",col=color.rf)
```